---
title: "Intro to ALTREP and Compact Sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro to ALTREP and Compact Sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## ALTREP 101

*Note: In all the `altrepr` vignettes, footnotes^[Like this one!] are used to point to relevant
information in the R source code. A casual reader need not follow any of these
links.*

ALTREP, short for alternate representation, is a mechanism where base R and 
package authors can write alternative versions of some of R's base data types,
with a different layout in memory.

There are many possible reasons why it might be advantageous to do this:

* Data can be represented more compactly (as discussed in this vignette)
* Functions can reduce computation by returning lazy ALTREP results that only evaluate entries if they are needed (see `vignette("deferred_string")`)
* Metadata can be added to vectors that is invisible and un-editable by users (see `vignette("wrapper_objects")`)
* The storage mechanism of a vector can be abstracted away, allowing data which might not even exist in memory (see `vignette("memmap")`)

At present, types that can be ALTREPPED include^[[altrep.c#L131-L137](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/altrep.c#L131-L137)]:

* Integer vectors
* Double vectors
* Logical vectors
* String vectors
* Complex vectors
* Raw vectors
* Lists

Some more advanced data types such as environments currently can't have ALTREPs.

Although we consider ALTREPs to have a different memory representation, in the
end they are actually stored using R's normal SEXPREC structures like lists, 
environments and vectors. The difference is that authors have the flexibility,
to, for example, use a list to represent a vector, or use several vectors to 
represent a single vector.

Currently some of the behaviours that can be overridden in ALTREP include^[[altrep.c#L139-L203](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/altrep.c#L139-L203)]:

* How the length is calculated (as returned by `length()`)
* How the data is converted to an array of data in C
* How elements are extracted from a vector in R (using `[` or `[[`)
* If the array is considered sorted
* How the minimum or maximum values of a vector are calculated

You might wonder why all of this couldn't be done with one of R's class systems,
like S3 or S4, since they support overriding the behaviour of `length`,
`[`, `max` etc. Some reasons include:

* ALTREP allows editing behaviours that cannot be edited in S3, for example 
  making a custom `DATAPTR`, the pointer to the vector data.
* Custom ALTREP state is hidden from the user (although `altrepr` makes this less true)
* Authors can use hacks that would break if exposed to R, such as using a character vector containing null pointers^[[altclasses.c#L688](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/altclasses.c#L688)]
* Presumably method dispatch is faster in ALTREP because it doesn't involve searching the S3 method table. 
  But this should be verified by benchmarks.

## Background

Let's get hands-on with  `altrepr` by looking at the "compact sequences",
perhaps the most famous ALTREP class in R.

Compact sequences [were introduced in R 3.5.0](https://cran-archive.r-project.org/bin/windows/base/old/3.5.0/NEWS.R-3.5.0.html).
They are simply integer (or double) vectors that are represented as a "range": they store the start, length, and step of a sequence, rather than storing every individual element in that range, which can save a lot of memory. 
Compact sequences are created by the `:` operator, and by the `seq` family of functions. 
Let's investigate!

## The Tip of the Iceberg

To start we need to install and load the `altrepr` package:

```{r, eval=FALSE}
remotes::install_github("multimeric/altrepr")
```
```{r setup}
library(altrepr)
```

Firstly, we can use the `is_altrep()` function to distinguish between ALTREP and 
ordinary vectors:

```{r}
is_altrep(c(1, 2, 3))
```
```{r}
is_altrep(1:3)
```
```{r}
is_altrep(seq(1, 3))
```

Next, we can get some details about an ALTREP's data using `alt_details()`:

```{r}
alt_details(1:3)
```

* `class_name` is the human-readable class name
* `pkg_name` is the package where the ALTREP was defined
* `base_type` is the R data type that the ALTREP is representing

We'll go into a bit more detail about the data fields later on.

If we want, instead of using `alt_details()`, we can used some more targeted
functions to access these same fields:

```{r}
alt_classname(1:3)
```

```{r}
alt_pkgname(1:3)
```

## Compact Real

There is actually another related class to `compact_intseq`, which you can
only get by converting an intseq to double:

```{r}
x <- as.double(1:5)
alt_details(x)
```

This has almost all of the same properties as `compact_intseq`, but of course
it is treated as a `numeric`/`real`/`double` by R.

## Function Naming Scheme

Functions in `altrepr` starting with `alt_` or `is_alt_` relate to any ALTREP class.
More specific ALTREP classes have their own utility functions. 
For compact vectors, this prefix is `compact_` and `is_compact`.

The first example we will see of this is a simple check for compact vectors:

```{r}
is_compact_vec(1:3)
```

```{r}
is_compact_vec(c(1, 2))
```

## Coercing to a Standard Integer Vector

The ALTREP API doesn't currently provide a function for forcing an ALTREP vector
to be converted to it's standard representation form, but compact vectors can be
expanded using any operation that clones/duplicates the vector.

The simplest way to copy a vector is to use an empty index on it (`[]`).
We can use this feature to prove the importance of the compact sequence class, 
by comparing the memory usage of a traditional integer vector, and the compact version:

Let's start with an enormous vector:
```{r}
x <- 1:10^9
is_altrep(x)
```

But even with 1 billion elements, it's actually not very big!
```{r}
lobstr::obj_size(x)
```

Now let's force the vector to expand. You will notice that this seemingly simple
operation actually takes suspiciously long. 
This is of course because a 1 billion element
standard representation vector is being generated behind the scenes:
```{r}
system.time({
  y <- x[]
})
```
```{r}
is_altrep(y)
```


```{r}
lobstr::obj_size(y)
```

There we go. It's no longer ALTREP, and we've worked out that we saved about 4 GB by 
using ALTREP!

For reference, `altrepr` also provides a utility function for this use case:
```{r}
y <- compact_to_standard(x)
is_altrep(y)
```
```{r}
lobstr::obj_size(y)
```

## Reading ALTREP Data

ALTREP classes have two "slots" for storing data (not the same as S4 slots,
but it's a helpful
analogy). These are respectively called `data1` and `data2`.
These slots can store any type of R object, including
recursive types like lists, so this isn't really a restriction. 

### `data1`

In the case of compact sequences, `data1` is used to store the parameters
of the sequence as a double (not integer!) vector

* The first entry contains the length of the sequence
* The second entry contains the start value of the sequence
* The third entry contains the step, which is currently always 1 or -1

We can prove all this using the `alt_data1()` function:

```{r}
alt_data1(1:3)
```

```{r}
alt_data1(2:3)
```

```{r}
alt_data1(3:2)
```

Actually `altrepr` has a utility function for finding this information, specifically for compact vectors:

```{r}
compact_details(4:2)
```


### `data2` and the Expanded Form
Compact seqs are considered to start in "compact" form, which we can see at the very end of the output from `alt_inspect()`. 
This function prints some internal information about the altrep vector which often seems to be informative when dealing with built-in ALTREP types:

```{r}
x <- 1:3
alt_inspect(x)
```

"compact" means that `data2` is `NULL`, which is its initial value:

```{r}
alt_data2(x)
```

A shortcut method to check for this is `compact_is_expanded`:

```{r}
compact_is_expanded(x)
```

When the `DATAPTR` of the sequence is accessed (which is a pointer to the array of data in the vector), it forces the compact sequence to expand.
`altrepr` has a special built-in function that forces a compact vector to expand without any other side effects:

```{r}
compact_expand(x)
alt_inspect(x)
```

```{r}
compact_is_expanded(x)
```


Notably `x` is *still* ALTREP, it hasn't been coerced into a standard representation vector:
```{r}
is_altrep(x)
```

But the `data2` value, which is linked to the expanded form, is now set to a full vector of integer data:

```{r}
alt_data2(x)
```

## Modifying ALTREP Data

**Note: improperly modifying ALTREP data can be potentially very dangerous and will risk crashing or corrupting your R session if done incorrectly!**.

`altrepr` also provides `set_alt_data1` and `set_alt_data2` for modifying the ALTREP data.
As an example, let's create a compact range and then modify it. 
This is safe because we're replacing a double vector with another double vector that 
follows the same layout as described above.


```{r}
x <- 1:5
set_alt_data1(x, c(2, 3, 1))
x
```

Firstly note that we've taken the range `1:5` and replaced it with `3:4`, because
we set the start of the range to be 3, and the length of it to be 2.

Also, importantly, note that **`x` has been modified in-place**.
This means that any shallow copy of the compact sequence 
(e.g. `y <- x` below)  will also be modified.
This means that there's actually no way to make a modified copy of `x` as in normal R:

```{r}
x <- 1:5
y <- x
set_alt_data1(x, c(2, 3, 1))
y
```

