---
title: "ALTREP Anatomy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{altrep_anatomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*This vignette goes into slightly more detail than the other vignettes
about memory representations and C code*.

You might be surprised to discover that an `ALTREP` is actually stored using
the same memory layout as an R pairlist, which is the same data structure used
to store formal arguments:

```{r}
paste0 |>
  formals() |>
  mode()
```

R pairlists hold 3 pointers, (called the `CAR`, `CDR` and `TAG`)^[[Defn.h#L145-L149](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/include/Defn.h#L145-L149)].
ALTREPs re-use this structure to point to the three parts of an ALTREP^[[Rinlinedfuns.h#L386-L389](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/include/Rinlinedfuns.h#L386-L389)]:

  * `CAR` -> data1: internal data, as a `SEXPREC` (normal R object)
  * `CDR` -> data2: more internal data (also a `SEXPREC`)
  * `TAG` -> class: The ALTREP class, as a raw vector with attributes
  
In addition, ALTREPs are distinguished from other R structures via the `ALTREP`
flag, which is set to 1^[[Defn.h#L221-L222](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/include/Defn.h#L221-L222) and [Defn.h#L118](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/include/Defn.h#L118)].
  
By now you are probably familiar with `alt_data1` and `alt_data2` for prying into those
first two fields.
But for the `TAG`, `altrepr` actually has another trick up its sleeve: `alt_class`.

As mentioned, the ALTREP class, not to be confused with the ALTREP class *name*, 
is a raw vector. 
We can look at it, but it's not particularly interesting:

```{r}
library(altrepr)
alt_class(1:3)
```

One potential utility of this is checking if two objects belong to the same
class:

```{r}
identical(
  alt_class(1:4),
  alt_class(100:90)
)
```
```{r}
identical(
  alt_class(1:4),
  sort(5:1)
)
```

R basically does the same thing here^[[altrep.c#L1160-L1163](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/altrep.c#L1160-L1163)].

Possibly more interesting is the *attributes* of the class, which is where we 
get all the juicy details:
```{r}
1:5 |>
  alt_class() |>
  attributes()
```

Notice that this is the source for many of the fields returned by `alt_details`:

```{r}
alt_details(1:5)
```


