---
title: "Intro to ALTREP and Compact Sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{compact_seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# ALTREP

ALTREP, short for alternate representation, is a mechanism where base R and 
package authors can write alternative versions of some of R's base data types. 

There are many possible reasons why it might be advantageous to do this:
* Data could be represented more compactly (as discussed in this vignette)
* Functions can reduce computation by returning lazy ALTREP results that only evaluate entries if they are needed (see [deferred_string.html])
* Adding metadata to vectors that is invisible and un-editable by users (see [wrapper_objects.html])
* Abstracting away the actual storage mechanism of a vector, which might not even exist in memory (see [memmap.html]) 

At present, types that can be ALTREPped include:
* Integer vectors
* Double vectors
* Logical vectors
* String vectors
* Complex vectors
* Raw vectors
* Lists
Some more advanced data types such as environments current can't have ALTREPs.

ALTREPs are stored using R's normal SEXP mechanism, but have some unique 
properties:

* The ALTREP struct holds three pointers:
  * data1: another `SEXP`
  * data2: another `SEXP`
  * tag: pointing to a raw vector that describes the ALTREP class
* The `ALTREP` flag is set to 1

Currently some of the behaviours that can be overridden in ALTREP include:
* How the length is calculated (as returned by `length()`)
* How the data is converted to an array of data in C
* How elements are extracted from a vector in R (using `[` or `[[`)
* If the array is considered sorted
* How the minimum or maximum values of a vector are calculated

You might wonder why all of this couldn't be done with one of R's class systems,
like S3 or S4, since they support overriding 

# Background

Let's learn about `altrepr` by looking at the "compact sequences", perhaps the most famous 
ALTREP class in R.

Compact sequences [were introduced in R 3.5.0](https://cran.r-project.org/bin/windows/base/old/3.5.0/NEWS.R-3.5.0.html).
They are simply integer (or double) vectors that are represented as a "range": they store the start, length, and step of a sequence, rather than storing every individual element in that range, which can save a lot of memory. Compact sequences are created by the `:` operator, and by the `seq` family of functions. Let's investigate this class using altrepr!

# The Tip of the Iceberg

To start we need to load the `altrepr` package:

```{r setup}
library(altrepr)
```

Firstly, we can use the `is_altrep` class to distinguish between ALTREP and 
ordinary vectors:

```{r}
is_altrep(c(1, 2, 3))
```
```{r}
is_altrep(1:3)
```
```{r}
is_altrep(seq(1, 3))
```

We can get the ALTREP classname using `alt_classname`:

```{r}
alt_classname(1:3)
```

We can find out where the ALTREP class was defined using `alt_pkgname`:

```{r}
alt_pkgname(1:3)
```

# Function Naming Scheme

Functions in `altrepr` starting with `alt_` or `is_alt_` relate to any ALTREP class.
More specific ALTREP classes have their own utility functions. 
For compact vectors, this prefix is `compact_`.

The first example we will see of this is a simple check for compact vectors:

```{r}
is_compact_vec(1:3)
```

```{r}
is_compact_vec(c(1, 2))
```

# Compact Real

There is actually another related class to `compact_intseq`, which you can
only get by converting an intseq to double:

```{r}
x <- as.double(1:5)
alt_classname(x)
```

This has almost all of the same properties as `compact_intseq`, but of course
it is treated as a `numeric`/`real`/`double` by R.

# Coercing to a Standard Integer Vector

The ALTREP API doesn't currently provide a function for forcing an ALTREP vector
to be converted to it's standard representation form, but compact vectors can be
expanded using any operation that clones/duplicates the vector.

The simplest way to copy a vector is to use an empty index on it (`[]`), although
note that this won't copy attributes, only vector values.

We can use this feature to prove the value of the compact sequence class, 
by comparing the memory usage of a traditional integer vector, and the compact version:

Let's start with an enormous vector:
```{r}
x <- 1:10^9
is_altrep(x)
```

But even with 100 million elements, it's actually not very big!
```{r}
lobstr::obj_size(x)
```

Now let's force the vector to expand. You will notice that this seemingly simple
operation actually takes suspiciously long. This is of course because a 1 billion element
standard representation vector is being generated behind the scenes:
```{r}
system.time({
  y <- x[]
})
```
```{r}
is_altrep(y)
```


```{r}
lobstr::obj_size(y)
```

There we go. It's no longer ALTREP, and we've worked out that we saved about 4 GB by 
using ALTREP!

For reference, `altrepr` also provides a utility function for this use case:
```{r}
y <- compact_to_standard(x)
is_altrep(y)
```
```{r}
lobstr::obj_size(y)
```

# Reading ALTREP Data

ALTREP classes have two "slots" for storing data (not the same as S4 slots, but it's a helpful
analogy). These are respectively called `data1` and `data2`.
These slots can store any type of R object, including
recursive types like lists, so this isn't really a restriction. 

## `data1`

In the case of compact sequences, `data1` is used to store the parameters
of the sequence as a double (not integer!) vector

* The first entry contains the length of the sequence
* The second entry contains the start value of the sequence
* The third entry contains the step, which is currently always 1 or -1

We can prove all this using the `alt_data1()` function:

```{r}
alt_data1(1:3)
```

```{r}
alt_data1(2:3)
```

```{r}
alt_data1(3:2)
```

Actually `altrepr` has a utility function for finding this information, specifically for compact vectors:

```{r}
compact_fields(4:2)
```


## `data2` and the Expanded Form
Compact seqs are considered to start in "compact" form, which we can see at the very end of the output from `alt_inspect`. 
This function prints some internal information about the altrep vector which often seems to be informative when dealing with built-in ALTREP types:

```{r}
x <- 1:3
alt_inspect(x)
```

"compact" means that `data2` is `NULL`, which is its initial value:

```{r}
alt_data2(x)
```

A shortcut method to check for this is `compact_is_expanded`:

```{r}
compact_is_expanded(x)
```

When the `DATAPTR` of the sequence is accessed (which is a pointer to the array of data in the vector), it forces the compact sequence to expand.
`altrepr` has a special built-in function that forces a compact vector to expand without any other side effects:

```{r}
compact_expand(x)
alt_inspect(x)
```

```{r}
compact_is_expanded(x)
```


Notably `x` is *still* ALTREP, it hasn't been coerced into a standard representation vector:
```{r}
is_altrep(x)
```

But the `data2` value, which is linked to the expanded form, is now set to a full vector of integer data:

```{r}
alt_data2(x)
```

## Modifying ALTREP Data

**Note: improperly modifying ALTREP data can be potentially very dangerous and will risk crashing or corrupting your R session if done incorrectly!**.

`altrepr` also provides `set_alt_data1` and `set_alt_data2` for modifying the ALTREP data.
As an example, let's create a compact range and then modify it. 
This is safe because we're replacing a double vector with another double vector that 
follows the same layout as described above.


```{r}
x <- 1:5
set_alt_data1(x, c(2, 3, 1))
x
```

Firstly note that we've taken the range `1:5` and replaced it with `3:4`, because
we set the start of the range to be 3, and the length of it to be 2.

Also, importantly, note that **`x` has been modified in-place**.
As noted above, any copy of a compact sequence will no longer be ALTREP.
This means that there's actually no way to make a modified copy of `x` as in normal R.
