---
title: "Deferred Strings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deferred Strings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Basics

*Note: you should have read `vignette("altrepr")` in order to understand this one*

Another built-in ALTREP class in R is the "Deferred String", a class that is instantiated when you convert a double to a character vector^[[coerce.c#L1290-L1298](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/coerce.c#L1290-L1298)]:

```{r}
library(altrepr)
x <- as.character(1)
is_altrep(x)
```
```{r}
alt_classname(x)
```

Or when you convert an integer vector to a character vector:

```{r}
1L |> as.character() |> alt_classname()
```

Interestingly the same is not the case for logical conversions:

```{r}
TRUE |> as.character() |> is_altrep()
```

# Memory Saving

The purpose of doing this deferred conversion is probably to save memory, 
since in most cases it's smaller to store the original numeric vector as numeric, 
and only convert entries to character on demand.

```{r}
x <- as.character(1:10)
lobstr::obj_size(x)
```
As with compact sequences, we can convert the ALTREP to standard rep using `[]`:

```{r}
y <- x[]
is_altrep(y)
```
```{r}
lobstr::obj_size(y)
```

At 10 entries, the deferred ALTREP representation is actually a bit larger than the standard representation, but anything much larger is vastly more efficient:

```{r}
x <- as.character(1:1000)
lobstr::obj_size(x)
```
```{r}
x[] |> lobstr::obj_size()
```

# Data

The first data slot for a deferred string is basically just a copy of the original numeric vector the deferring string was initialised using^[[altclasses.c#L890-L891](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/altclasses.c#L890-L891)].
Unfortunately for some reason (a bug?), we can't actually interrogate this vector directly, or it will crash the R interpreter.

At best we can inspect the deferred string:

```{r}
x <- as.character(1:5)
alt_inspect(x)
```

The top line is showing that `x` itself is a deferred string, and the second line
indicates that it contains an integer vector (`INTSXP`) as explained above.

`data2` is a bit easier to interrogate. 
It's simply a character vector that caches the "true" result of the conversion as a character vector.

```{r}
alt_data2(x)
```

According to the C code, each entry of `data2` is set to `NULL` initially, 
and each is "expanded" to their final value on-demand, as they are needed^[[altclasses.c#L689-L695](https://github.com/wch/r-source/blob/7d5117ad947beed213d3d4538bfe36441478ebef/src/main/altclasses.c#L689-L695)].
However this is difficult to demonstrate because accessing single elements doesn't
seem to actually expand them as we might expect.

However, it's easy to see how changing the character vector in `data2` affects
the calculated elements of the array:

```{r}
set_alt_data2(x, LETTERS[1:5])
x
```

# Compact and Expanded

Like compact sequences, deferred strings are considered to be either compact or expanded.
We can expand the string and check its state using corresponding utility methods:

```{r}
x <- as.character(1:5)
deferred_is_expanded(x)
```

```{r}
deferred_expand(x)
deferred_is_expanded(x)
```

However, unlike compact sequences, the amount of data in `data2` isn't the deciding factor in
a sequence being expanded. It's actually the *absence* of data in `data1` that does it.

```{r}
x <- as.character(1:5)
set_alt_data1(x, NULL)
alt_inspect(x)
```



