---
title: "memmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{memmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Memory mapping (memmap) refers to mapping a file on disk to an object in memory.
Data is only loaded from disk as required, meaning that the entire object
can be much larger than there is space for in memory.

Apparently, R contains a family of memmap ALTREP classes that provide this behaviour.

# Creating a Memory Map

As far as I can tell, the only public interface for the memmap classes provided by R is
another internal function.

This has 5 possible arguments. The important ones are:

1. The file to memory map, as a character scalar (required)
2. The vector type to map it to, either "double" or "integer" (optional, defaults to "double")

To test this, we first create a dataset to play with:

```{r}
data <- rnorm(n=5)
head(data)
```

Then we write this dataset to a temporary file:

```{r}
filename <- tempfile()
writeBin(data, filename)
```

Finally, we can access it again as a memory mapped vector **that exists on disk!**
```{r}
mapped <- .Internal(mmap_file(filename))
mapped
```


# Prying into the Memory Map

The classname is `mmap_` + data type, which in this case is `real`:

```{r}
library(altrepr)
alt_classname(mapped)
```

The `data1` slot is a pointer to the memory map itself, in memory (although it's
a kind of virtual pointer because the memory exists on disk):

```{r}
alt_data1(mapped)
```

The `data2` slot contains a list, whose entries are ^[https://github.com/wch/r-source/blob/a0409ebcd1fc16247812d425085c8a3cc4aa4499/src/main/altclasses.c#L927-L949]:

1. The name of the file we have memory mapped
2. A length-2 vector:
    1. The size of the memory map in bytes
    2. The number of entries in the vector
3. A length-4 vector:
    1. The `SEXPTYPE` of the memory map (either 13 for integer, or 14 for double)
    2. A boolean called `ptrOK`. Not sure what this does
    3. A boolean, which is `1` if the memory map is requested to be writable
    4. Another boolean called `serOK`. Not sure what this means either.

```{r}
alt_data2(mapped)
```


