---
title: "Memory Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Memory Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

*Note: you should have read `vignette("altrepr")` in order to understand this one*

Memory mapping (memmap) refers to mapping a file on disk to an object in memory.
Data is only loaded from disk as required, meaning that the entire object
can be much larger than there is space for in memory.

Apparently, R contains a family of memmap ALTREP classes that provide this behaviour.

# Creating a Memory Map

Seemingly, the only interface for the memmap classes provided by R is
another internal function.
In other words, memmap classes have no public constructor in R yet.

That doesn't mean that `altrepr` doesn't help, though.
The `mmap_make()` function attempts to turn this internal constructor into
something a bit more user friendly.

To test it, we first create a dataset to play with:

```{r}
data <- rnorm(n=5)
data
```

Then we write this dataset to a temporary file, and open it as a memory mapped vector **that exists on disk!**

```{r}
library(altrepr)
filename <- tempfile()
mapped <- mmap_make(filename, data)
mapped
```

# Prying into the Memory Map

The classname is `mmap_` + data type, which in this case is `real`:

```{r}
alt_classname(mapped)
```

The `data1` slot is a pointer to the memory map itself, in memory (although it's
a kind of virtual pointer because the memory exists on disk):

```{r}
alt_data1(mapped)
```

The `data2` slot contains a list, whose entries are^[[altclasses.c#L927-L949](https://github.com/wch/r-source/blob/a0409ebcd1fc16247812d425085c8a3cc4aa4499/src/main/altclasses.c#L927-L949)]:

1. The name of the file we have memory mapped
2. A length-2 vector:
    1. The size of the memory map in bytes
    2. The number of entries in the vector
3. A length-4 vector:
    1. The `SEXPTYPE` of the memory map (either 13 for integer, or 14 for double)
    2. A boolean called `ptrOK`. It is unclear what this does or was planned to do.
    3. A boolean, which is `1` if the memory map is requested to be writable
    4. Another boolean called `serOK`. Also unclear what this means.

```{r}
alt_data2(mapped)
```

As usual, `altrepr` provides a utility function for more easily accessing these fields:

```{r}
mmap_details(mapped)
```
