---
title: "Wrapper Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wrapper Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

*Note: you should have read `vignette("altrepr")` in order to understand this one*

Wrapper objects have a (rare) description in the R source code^[[altclasses.c#L1423-L1424](https://github.com/wch/r-source/blob/a0409ebcd1fc16247812d425085c8a3cc4aa4499/src/main/altclasses.c#L1423-L1424)]. Good start!

> Wrapper objects are ALTREP objects designed to hold the attributes
   of a potentially large object and/or meta data for the object.
   
# Metadata
   
Actually the metadata that's being referred to here is not something obvious
like the `attributes()` of an object. It's simply a vector of integers that
describe a vector in some way^[[altclasses.c#L1431-L1432](https://github.com/wch/r-source/blob/a0409ebcd1fc16247812d425085c8a3cc4aa4499/src/main/altclasses.c#L1431-L1432)]. Currently these include:

1. Sortedness. Describes the sorted state of a vector^[[Rinternals.h#L1144-L1164](https://github.com/wch/r-source/blob/a0409ebcd1fc16247812d425085c8a3cc4aa4499/src/include/Rinternals.h#L1144-L1164)]. Can be:
    * `SORTED_DECR_NA_1ST`: -2 (sorted in decreasing order, with `NA` values sorted first)
    * `SORTED_DECR`: -1 (sorted in decreasing order)
    * `KNOWN_UNSORTED`: 0
    * `SORTED_INCR`: 1 (sorted in increasing order)
    * `SORTED_INCR_NA_1ST`: 2
    * `UNKNOWN_SORTEDNESS`: `NA_integer_`
2. No `NA`. `0` if a vector might possibly contain `NA` values, otherwise `1`.

The public API in C for this are functions such as `INTEGER_NO_NA`, `INTEGER_IS_SORTED` etc^[[Rinternals.h#L295-L303](https://github.com/wch/r-source/blob/a0409ebcd1fc16247812d425085c8a3cc4aa4499/src/include/Rinternals.h#L295-L303)].

# Wrapper Objects in the Wild

The most likely situation you will encounter wrapper objects in normal R usage is via the `sort` function.
Most (but not all) parameter configurations of `sort` will cause the function to return an ALTREP:

```{r}
library(altrepr)
x <- sort(5:1)
is_altrep(x)
```

We can do our normal prying into the class metadata:

```{r}
alt_classname(x)
```
```{r}
alt_pkgname(x)
```

# Inspecting Wrapper Metadata in R

The main public interface to the sorted metadata is the `is.unsorted` function,
which basically checks the metadata using mechanisms discussed above.
This is a "true public" function in the sense that it's a stable function provided
by base R, that you can rely on in the long term:

```{r}
is.unsorted(5:1)
```

```{r}
5:1 |> sort() |> is.unsorted()
```

The main public interface to the NA flag is the [anyNA()](NA) function:

```{r}
c(1, 2, NA, 3) |> anyNA()
```
```{r}
c(1, 2, 3) |> anyNA()
```

Of course `altrepr` provides a user-friendly utility for inspecting the metadata of known
wrapper objects:

```{r}
5:1 |>
  sort() |>
  wrapper_details()
```
# Creating Custom Wrapper Types

`altrepr` also provides a user-friendly function for creating these wrapper types, without needing `sort`.
This lets us blatantly lie to R about our vectors!

```{r}
wrapper <- wrapper_make(c(5, 3, NA, 1), is_sorted = TRUE, has_na = FALSE)
wrapper
```
```{r}
wrapper_details(wrapper)
```

This fools the public R functions as well:

```{r}
is.unsorted(wrapper)
```
```{r}
anyNA(wrapper)
```

There's also an internal R function you can use (at your own risk) called `wrap_meta`.
This isn't documented, but it takes 3 arguments:

1. The R object to wrap
2. An integer indicating the sortedness (see above for the sortedness enum values)
3. An integer indicating whether it contains `NA`. `1` if it does, otherwise `0`.

```{r}
x <- wrap_meta(10:1, 0, 0) |> .Internal()
alt_classname(x)
```
```{r}
wrap_meta(10:1, 1, 0) |> 
  .Internal() |>
  is.unsorted()
```

# Wrapper Data

The data for these types is quite simple. 
`data1` contains the wrapped object, and `data2` contains the metadata as
an integer vector. Let's have a look:

```{r}
x <- sort(5:1)
alt_data1(x)
```
```{r}
alt_data2(x)
```

Remember that 1 1 means "sorted in increasing order" and "no NA values".
